Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-09-14 Thesis/Background.tex
Most corrections implemented; expanding discussions tomorrow

\chapter{A Dynamical Recipe for Cosmological Disks}\label{ch:background}
\newpage

This chapter provides an overview of the background theory needed to understand subsequent chapters. In Chapter~\ref{ch:introduction}, we discussed simulations of galaxies being critical to having a theoretical understanding of cosmology and galaxy formation. The word ``simulation" was used without fully contextualizing its significance and meaning. We provide that context in \S\ref{sec:motivation}, where we describe what is actually being done when a simulation is run. In \S\ref{sec:galaxy_ics}, we talk about how to apply this theory to study the evolution of equilibrium galaxies.  We talk about  the implications of cosmology in detail in \S\ref{sec:cosmology}, and explain how this specifies an extended view of the discussion in \S\ref{sec:motivation}. This includes how we identify cosmic substructure and the techniques we use for analyzing the time-evolution of disk-derived quantities. 
\section{Physical Motivation of Modelling} \label{sec:motivation}

@@

We have shown how to map solving the CBE, the master equation for our system of stars and dark matter, to the evolution of a finite sample of particles. The system of equations we need to solve now is,
\begin{eqnarray} \label{eq:nbody_system}
\dot{\textbf{r}}_i(t) &=& \textbf{v}_i(t)\\
\dot{\textbf{v}}_i(t) &=& -\left.\deriv{\Phi}{\textbf{r}}\right\vert_{\textbf{r} = \textbf{r}_i}.
\end{eqnarray}
where $\dot{\textbf{r}_i}$ and $\dot{\textbf{v}}_i$ are the first time derivatives of $\textbf{r}_i$ and $\textbf{v}_i$, respectively. Our integro-differential equation system is now approximated as $6\,N$, where $N$ is the number of particles sampled to represent the stars and dark matter, ordinary differential equations (ODEs). These may be solved by any ODE system solver. There are two competing objectives when deciding on the best way to handle time stepping numerically; the first is that we want to maximize the accuracy of our integration scheme and the second is that we want to minimize the number of evaluations of $-\nabla \Phi$, the force. The latter consideration is a severe constraint because naively, the complexity of evaluating each pairwise force is $O(N^2)$.  Although, as we will see, approximations reduce this complexity to $O(N\,\log \, N)$. We note that finding the forces amounts to solving \eqref{eq:poisson}.

@@

The level of temporal refinement assigned to a given particle is determined primarily by the acceleration it experiences. That is, for \textsc{Gadget-3} \citep{GadgetCodePaper},
\begin{equation}
\Delta t_i = \min\left(\Delta t_{base}, \left(\frac{2 \eta \epsilon}{\vert \nabla \Phi(\textbf{r}_i) \vert} \right)^{1/2}\right).
\end{equation}
Here, $\eta$ is a free accuracy parameter and $\epsilon$ is the gravitational softening. The gravitational softening is used to prevent numerical overflow when particles rarely have close encounters. It arises from having the acceleration induced from a particle at position $\textbf{r}_i$  be,
\begin{equation}
\textbf{a}_i(\textbf{r}) = -\frac{G m_i}{\left(\vert\textbf{r} - \textbf{r}_i\vert^2 + \epsilon^2\right)^{3/2}}(\textbf{r} - \textbf{r}_i) \label{eq:point_mass}
\end{equation}
Higher choices of $\epsilon$ reduce large errors accumulating due to large forces, but also make the force calculation less accurate.

@@

To summarize, there are a number of ways to find gravitational forces in a simulation that all amount to solving Poisson's Equation, \eqref{eq:poisson}. The most accurate way is to compute all pairwise forces, but efficient algorithms have been developed to accurately estimate forces from gravity. The force calculation together with a time stepping scheme for N-body realizations forms the main components of an N-body simulation, and solving Eq. \eqref{eq:nbody_system} to estimate $f(\textbf{r},\textbf{v},t)$ is what we mean when we say we ran a simulation.

@@

\section{Phase Space and Initial Conditions} \label{sec:galaxy_ics}
In \S\ref{sec:motivation} we outlined the underlying theory behind representing stars and dark matter in an N-body simulation. The focus of this thesis is studying the behavior of stellar disks in cosmological environments, and the first step towards that understanding is the construction of a pristine, equilibrium N-body model. Ultimately, in the course of running an N-body simulation, these pristine systems will diverge to higher entropy\footnote{The Shannon entropy is a commonly used measure for self-gravitating systems. See \citet{BT} for more details.} states, initially due to random errors in the N-body representation. Nonetheless, these models give us an understanding of how perfectly unperturbed galaxies will evolve due to the fact that real galaxies are also made of a finite number of particles.

@@

The DF represents the 6D phase space information and completely describes an N-body system. In a simple disk-halo system, there is a flattened axisymmetric component (the disk), and a spheroidal component (the dark matter halo) each with their own DF. These are made self-consistent by the separate DFs' incorporation of their combined gravitational potential.

However, the DF is not the quantity that is observed in external galaxies. Astronomers are not typically working with a 6D phase space description of the structures they observe. Luminosity (density) profiles are one observable , as well as circular velocities (rotation curves) for disks, and also the spread of line of sight (LOS) velocities of disk stars. A modelling approach that takes our understanding of these observables as input is conceptually easier to understand than a DF which may or may not be unique \citep{BT}.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} presented one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, disks, and dark matter halos. In the case of the dark matter halo for a disk-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} proceeded this work in establishing a common density profile for dark matter halos). As such, we will not talk about their uninformed choice of halo, and it really is not important.

In the case of the axisymmetric disk, \citet{hernquist_1993} based their model on the observations at the time which suggested that stellar disks had an exponential radial density profile \citep{freeman_1970}. This motivated a density profile commonly used today,
\begin{equation}
\rho_d(r) = \frac{M_d}{4 \pi R_d^2 z_d} e^{-R/R_d} \text{sech}^2\left(\frac{z}{z_d} \right). 
\end{equation}
Here, $M_d$ is the mass of the disk, $R_d$ is the radial scale length, and $z_d$ is the vertical scale length.
Observations support the idea that the second radial velocity moment also has an exponential profile \citep{van_der_kruit_searle_1981, lewis_freeman_1989},
\begin{equation}
\langle v_R^2 \rangle = \sigma_{R,0} e^{-R/R_\sigma}.
\end{equation}
Here, $R_\sigma$ is the radial dispersion scale length, and $\sigma_{R,0}$ is a free velocity scale (on the order of 100 km/s for the Milky Way). In  principle, any radial scale length can be used for $\langle v_R^2 \rangle$, but it is generally accepted to be a longer scale length than $R_d$. The vertical dispersion and second velocity moment are determined from the assumption of the disc being isothermal,
\begin{equation}
\langle v_z^2 \rangle = \pi G \Sigma z_d
\end{equation}
where $\Sigma_d$ is the surface density as a function of radius. Finally, a relation between $\langle v_\phi^2 \rangle$ and the other second moments is needed to fully solve the second order system. Jeans modelling for discs injects the intuition behind a theoretical development known as the \textit{epicycle approximation}.

@@

If we assume the epicycle approximation, then $I_3 \approx E_z = \Phi(R,0) - \Phi(R,z) + \frac{1}{2} v^2$, where $v$ is the velocity of an orbit \citep{kd95galactics}. The vertical energy, $E_z$, is conserved so long as the disc is razor thin. Jeans showed that a DF that is a function of the integrals of motion describes an equilibrium distribution. More specifically, quoting from \citet{BT},
\begin{theorem}[The Jeans Theorem]
Any steady-state solution of the collisionless Boltzmann equation depends on the phase-space coordinates only through integrals of motion in a given potential, and any function of the integrals yields a steady-state solution of the collisionless Boltzmann equation.
\end{theorem}
Proof of this theorem can be obtained in \citet{jeans_1915} and in Chapter~4.2 of \citet{BT}, although it essentially amounts to observing that the time derivative of a function of time-independent quantities is zero by the chain rule. Now that we know that any DF of the integrals of motion gives a steady state solution, and that using the epicyclic approximation gives us three known integrals of motion, we can ask ourselves what kinds of DFs describe the NFW haloes and exponential disks that are easily obtained from Jeans modelling.

@@

For the tilde functions, we choose them to yield the closest thing to an exponential-sech$^2$ density profile with and exponential radial velocity dispersion as possible. Assuming an exponential radial dispersion profile, a disc density is obtained by iteratively adjusting $\tilde{\sigma_z}$ and $\tilde{\rho}$ such that \citep{kd95galactics},
\begin{equation}
\rho_d(R,z) = \frac{M_d}{8 \pi R_d^2 z_d} e^{-R/R_d} \text{erfc }\left(\frac{R - R_{out}}{\sqrt{2} \delta R_{out}}\right) \exp \left[ -0.8676 \frac{\Psi_z(R,z)}{\Psi_z(R,z_d)}\right]
\end{equation}
Here, $M_d$ is a parameter that is approximately equal to the mass of the disk, $R_{out}$ is a truncation radius, $\delta R_{out}$ is a parameter determining the sharpness of the truncation, and $\Psi_z$ is the negative vertical potential, $\Phi(R,0) - \Phi(R,z)$. The motivation behind the last term in the product is to yield a dropoff of vertical density close to $\text{sech}^2(z/z_d)$.

@@

Turning the definition of the actions into a transformation from a physically motivated conjugate coordinate system is difficult in general. Closed form solutions for axisymmetric systems exist if the potential if of the St\"ackel form\footnote{For more detail, see Chapter~3.5.3 of \citet{BT}.}, but in general, these transformations do not exist. A commonly used technique is to locally approximate the potential as being of the St\"ackel form \citet{binney_2012}. In practice, one is reduced to approximating the potential instead of the third integral of motion as in the case of \textsc{GalactICS}. For nearly planar disc systems, the differences are insignificant, but become more relevant as one wants to initialize a thicker disc \citep{vasiliev_2018,bauer2018b}.

@@

Following up on this concept, we introduce the notion of \textit{redshift}. The term derives from the fact that as objects become more distant, we expect the hubble flow to cause the emitters to move away from the observer. Photons emitted by a stationary object with a wavelength, $\lambda$, are observed with wavelength $a(t_{emit}) \lambda$, and we define the redshift, $z$,
\begin{equation}
1 + z = \frac{1}{a}
\end{equation}
in analogy to this phenomenon. This is a descending unit of time that is often used interchangeably with the scale factor, $a$.
\subsection{FRW Cosmology: Cosmological ICs}

>>>>

\chapter{A Dynamical Recipe for Cosmological Discs}\label{ch:background}
\newpage

This chapter provides an overview of the background theory needed to understand subsequent chapters. In Chapter~\ref{ch:introduction}, we discussed simulations of galaxies being critical to having a theoretical understanding of cosmology and galaxy formation. The word ``simulation" was used without fully contextualizing its significance and meaning. We provide that context in \S\ref{sec:motivation}, where we describe what is actually being done when a simulation is run. In \S\ref{sec:galaxy_ics}, we talk about how to apply this theory to study the evolution of equilibrium galaxies.  We talk about  the implications of cosmology in detail in \S\ref{sec:cosmology}, and explain how this specifies an extended view of the discussion in \S\ref{sec:motivation}. This includes how we identify cosmic substructure and the techniques we use for analyzing the time-evolution of disc-derived quantities. 
\section{Physical Motivation of Modelling} \label{sec:motivation}

@@

We have shown how to map solving the CBE, the master equation for our system of stars and dark matter, to the evolution of a finite sample of particles. The system of equations we need to solve now is,
\begin{eqnarray} 
\dot{\textbf{r}}_i(t) &=& \textbf{v}_i(t)\label{eq:nbody_system_1}\\
\dot{\textbf{v}}_i(t) &=& -\left.\deriv{\Phi}{\textbf{r}}\right\vert_{\textbf{r} = \textbf{r}_i}.\label{eq:nbody_system_2}
\end{eqnarray}
where $\dot{\textbf{r}_i}$ and $\dot{\textbf{v}}_i$ are the first time derivatives of $\textbf{r}_i$ and $\textbf{v}_i$, respectively. Our integro-differential equation system is now approximated as $6\,N$, where $N$ is the number of particles sampled to represent the stars and dark matter, ordinary differential equations (ODEs). These may be solved by any ODE system solver. There are two competing objectives when deciding on the best way to handle time stepping numerically; the first is that we want to maximize the accuracy of our integration scheme and the second is that we want to minimize the number of evaluations of $-\nabla \Phi$, the force. The latter consideration is a severe constraint because naively, the complexity of evaluating each pairwise force is $O(N^2)$.  Although, as we will see, approximations reduce this complexity to $O(N\,\log \, N)$. We note that finding the forces amounts to solving \eqref{eq:poisson}.

@@

The level of temporal refinement assigned to a given particle is determined primarily by the acceleration it experiences. That is, for \textsc{Gadget-3} \citep{GadgetCodePaper},
\begin{equation}
\Delta t_i = \min\left(\Delta t_{base}, \left(\frac{2 \eta \epsilon}{\vert \nabla \Phi(\textbf{r}_i) \vert} \right)^{1/2}\right).
\end{equation}
Here, $\eta$ is a free accuracy parameter and $\epsilon$ is the gravitational softening. The gravitational softening is used to prevent numerical overflow when particles rarely have close encounters. It arises from having the acceleration induced from a particle at position $\textbf{r}_i$  be,
\begin{equation}
\textbf{a}_i(\textbf{r}) = -\frac{G m_i}{\left(\vert\textbf{r} - \textbf{r}_i\vert^2 + \epsilon^2\right)^{3/2}}(\textbf{r} - \textbf{r}_i) \label{eq:point_mass}
\end{equation}
Larger choices of $\epsilon$ reduce large errors accumulating due to large forces, but also make the force calculation less accurate.

@@

To summarize, there are a number of ways to find gravitational forces in a simulation that all amount to solving Poisson's Equation, \eqref{eq:poisson}. The most accurate way is to compute all pairwise forces, but efficient algorithms have been developed to accurately estimate forces from gravity. The force calculation together with a time stepping scheme for N-body realizations forms the main components of an N-body simulation, and solving Eqs. \eqref{eq:nbody_system_1} and \eqref{eq:nbody_system_2} to estimate $f(\textbf{r},\textbf{v},t)$ is what we mean when we say we ran a simulation.

@@

\section{Phase Space and Initial Conditions} \label{sec:galaxy_ics}
In \S\ref{sec:motivation} we outlined the underlying theory behind representing stars and dark matter in an N-body simulation. The focus of this thesis is studying the behavior of stellar discs in cosmological environments, and the first step towards that understanding is the construction of a pristine, equilibrium N-body model. Ultimately, in the course of running an N-body simulation, these pristine systems will diverge to higher entropy\footnote{The Shannon entropy is a commonly used measure for self-gravitating systems. See \citet{BT} for more details.} states, initially due to random errors in the N-body representation. Nonetheless, these models give us an understanding of how perfectly unperturbed galaxies will evolve due to the fact that real galaxies are also made of a finite number of particles.

@@

The DF represents the 6D phase space information and completely describes an N-body system. In a simple disc-halo system, there is a flattened axisymmetric component (the disc), and a spheroidal component (the dark matter halo) each with their own DF. These are made self-consistent by the separate DFs' incorporation of their combined gravitational potential.

However, the DF is not the quantity that is observed in external galaxies. Astronomers are not typically working with a 6D phase space description of the structures they observe. Luminosity (density) profiles are one observable , as well as circular velocities (rotation curves) for discs, and also the spread of line of sight (LOS) velocities of disc stars. A modelling approach that takes our understanding of these observables as input is conceptually easier to understand than a DF which may or may not be unique \citep{BT}.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} presented one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, discs, and dark matter halos. In the case of the dark matter halo for a disc-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} followed this work in establishing a common density profile for dark matter halos). As such, we will not talk about their uninformed choice of halo, and it really is not important.

In the case of the axisymmetric disc, \citet{hernquist_1993} based their model on the observations at the time which suggested that stellar discs had an exponential radial density profile \citep{freeman_1970}. This motivated a density profile commonly used today,
\begin{equation}
\rho_d(r) = \frac{M_d}{4 \pi R_d^2 z_d} e^{-R/R_d} \text{sech}^2\left(\frac{z}{z_d} \right). 
\end{equation}
Here, $M_d$ is the mass of the disc, $R_d$ is the radial scale length, and $z_d$ is the vertical scale length.
Observations support the idea that the second radial velocity moment also has an exponential profile \citep{van_der_kruit_searle_1981, lewis_freeman_1989},
\begin{equation}
\langle v_R^2 \rangle = \sigma_{R,0} e^{-R/R_\sigma}.
\end{equation}
Here, $R_\sigma$ is the radial dispersion scale length, and $\sigma_{R,0}$ is a free velocity scale (on the order of 100 km/s for the Milky Way). In  principle, any radial scale length can be used for $\langle v_R^2 \rangle$, but it is generally accepted to be a longer scale length than $R_d$. The vertical dispersion and second velocity moment are determined from the assumption of the disc being isothermal,
\begin{equation}
\langle v_z^2 \rangle = \pi G \Sigma z_d
\end{equation}
where $\Sigma_d$ is the surface density as a function of radius. Finally, a relation between $\langle v_\phi^2 \rangle$ and the other second moments is needed to fully solve the second order system. Jeans modelling for discs injects the intuition behind a theoretical development known as the \textit{epicycle approximation}.

@@

If we assume the epicycle approximation, then $I_3 \approx E_z = \Phi(R,0) - \Phi(R,z) + \frac{1}{2} v^2$, where $v$ is the velocity of an orbit \citep{kd95galactics}. The vertical energy, $E_z$, is conserved so long as the disc is razor thin. Jeans showed that a DF that is a function of the integrals of motion describes an equilibrium distribution. More specifically, quoting from \citet{BT},
\begin{theorem}[The Jeans Theorem]
Any steady-state solution of the collisionless Boltzmann equation depends on the phase-space coordinates only through integrals of motion in a given potential, and any function of the integrals yields a steady-state solution of the collisionless Boltzmann equation.
\end{theorem}
Proof of this theorem can be obtained in \citet{jeans_1915} and in Chapter~4.2 of \citet{BT}, although it essentially amounts to observing that the time derivative of a function of time-independent quantities is zero by the chain rule. Now that we know that any DF of the integrals of motion gives a steady state solution, and that using the epicyclic approximation gives us three known integrals of motion, we can ask ourselves what kinds of DFs describe the NFW haloes and exponential discs that are easily obtained from Jeans modelling.

@@

For the tilde functions, we choose them to yield the closest thing to an exponential-sech$^2$ density profile with and exponential radial velocity dispersion as possible. Assuming an exponential radial dispersion profile, a disc density is obtained by iteratively adjusting $\tilde{\sigma_z}$ and $\tilde{\rho}$ such that \citep{kd95galactics},
\begin{equation}
\rho_d(R,z) = \frac{M_d}{8 \pi R_d^2 z_d} e^{-R/R_d} \text{erfc }\left(\frac{R - R_{out}}{\sqrt{2} \delta R_{out}}\right) \exp \left[ -0.8676 \frac{\Psi_z(R,z)}{\Psi_z(R,z_d)}\right]
\end{equation}
Here, $M_d$ is a parameter that is approximately equal to the mass of the disc, $R_{out}$ is a truncation radius, $\delta R_{out}$ is a parameter determining the sharpness of the truncation, and $\Psi_z$ is the negative vertical potential, $\Phi(R,0) - \Phi(R,z)$. The motivation behind the last term in the product is to yield a dropoff of vertical density close to $\text{sech}^2(z/z_d)$.

@@

Turning the definition of the actions into a transformation from a physically motivated conjugate coordinate system is difficult in general. Closed form solutions for axisymmetric systems exist if the potential is of the St\"ackel from\footnote{For more detail, see Chapter~3.5.3 of \citet{BT}.}, but in general, these transformations do not exist. A commonly used technique is to locally approximate the potential as being of the St\"ackel form \citet{binney_2012}. In practice, one is reduced to approximating the potential instead of the third integral of motion as in the case of \textsc{GalactICS}. For nearly planar disc systems, the differences are insignificant, but become more relevant as one wants to initialize a thicker disc \citep{vasiliev_2018,bauer2018b}.

@@

Following up on this concept, we introduce the notion of \textit{redshift}. The term derives from the fact that as objects become more distant, we expect the Hubble flow to cause the emitters to move away from the observer. Photons emitted by a stationary object with a wavelength, $\lambda$, are observed with wavelength $a(t_{emit}) \lambda$, and we define the redshift, $z$,
\begin{equation}
1 + z = \frac{1}{a}
\end{equation}
in analogy to this phenomenon. This is a decreasing unit of time that is often used interchangeably with the scale factor, $a$.
\subsection{FRW Cosmology: Cosmological ICs}
